Light and growth form interact to shape stomatal ratio among British angiosperms

### 2017-10-14 ms/ms.Rnw
Added analysis of phylogenetic variance

I compared phylogenetic linear models to test whether growth form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$. Unlike the analysis above, there was significant phylogenetic signal in this comparison (see R code). I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface. I fit models using \pkg{phylolm} and calculated Akaike Information Criteria (AIC). For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution. I used a 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}. % REMOVING, but keep text in case I decide to replace: Likewise, to determine whether the interaction between Raunki\ae r life form and \el~was statistically significant, I used a parametric bootstrap to generate the null distribution of $\Delta$AIC values ($\Delta$AIC is the difference in AIC between competing models). Specifically, I sampled 1000 random datasets from the estimated model with main effects of Raunki\ae r life form and \el~but no interaction. I fit these simulated datasets to models with and without interactions and  calculated $\Delta$AIC. The statistical significance of the observed $\Delta$AIC is the proportion of simulated $\Delta$AIC greater than the observed.

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}). When classified based on plant habit, growth form interacted with \el less ($\Delta \textrm{AIC} = \Sexpr{round(AIC(fitSR_hf_g) - AIC(fitSR_hf_f), 1)}$; Fig.~\ref{fig:SR-hf}). Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore we focus hereafter on those analyses. Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SR-lf}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. % Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).

@@

% Revise and put after 2nd sentence: Although I repeated certain analyses from previous papers, this is an important advance because previous analyses did not include phylogeny or did so improperly... 
Previously, associations between light, growth form, and stomatal ratio have been interpreted in isolation as indicating that either high light and/or herbaceous growth form favors amphistomy. In British angiosperms, both factors are important, though light may be a stronger determinant than growth form based on information criterion comparisons (Table~\ref{table:srAIC}). Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesis (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs). This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}. The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g., containing both annual and perennial forms). I predict that amphistomy will be favored more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light, and much more strongly than a perennial grown in low light. Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991, Fontana_etal_2017}.

@@

The weak or nonsignificant relationship between \el~and stomatal ratio in geophytes and phanerophytes suggests that in some cases amphistomy may not reliably indicate open habitat without further information. For example, perhaps amphistomatous geophytes from partially shaded habitats are spring ephemerals, so they  experience high light during their growth phase, but this has not been tested. Likewise, phanerophytes (most tall trees) are almost always hypostomatous (see also \cite{Muir_2015}). Most British phanerophytes are tall, hypostomatous trees, but the exceptions are telling. For example, the most amphistomatous phanerophyte in this dataset is \textit{Brassica oleracea}, a short-statured biennial that has more in common physiologically with hemicryptophytes than other phanerophytes. The other amphistomatous phanerophytes in this data set (\textit{Populus nigra} and \textit{Lavatera arborea}) are fast-growing pioneer species.

>>>>

I compared phylogenetic linear models to test whether growth form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$. Unlike the analysis above, there was significant phylogenetic signal in this comparison (see R code). I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface. I fit models using \pkg{phylolm} and calculated Akaike Information Criteria (AIC). For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution. I used a 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}.

I also tested whether phylogenetic could explain the residual variation in stomatal ratio after accounting for growth form and \el. Specifically, I compared the expected residual variation given the actual tree versus a hypothetical tree where trait evolution has reached stationarity (i.e. a star phylogeny with infinite branch lengths). If phylogeny explains much of the variation, then the simulated residual variance from the actual tree should be greater than that of the stationary tree. I simulated trait values from 10$^4$ parametric bootstrap samples of the model with the lowest AIC (this was the model including Raunki\ae r lifeform, \el, and their interaction; see Results). I performed the first set of simulations using the actual phylogenetic tree in \pkg{OUwie} version 1.50 \citep{Beaulieu_OMeara_2016}. Each simulation used a different bootstrap parameter sample of $\alpha$ and $\sigma ^ 2$, where $\alpha$ is the return rate to the mean and $\sigma ^ 2$ is the diffusion rate. At stationarity, the variance of an Ornstein-Uhlenbeck trait is equal to $\sigma ^ 2 / 2 \alpha$. Therefore, I simulated stationary data by assuming a normal distribution with this variance estimated from the bootstrap samples. For comparability, I set the mean of simulations from both actual phylogeny and the stationary to zero. I compared the actual to stationary variance across simulated datasets using a paired $t$-test.

% REMOVING, but keep text in case I decide to replace: Likewise, to determine whether the interaction between Raunki\ae r life form and \el~was statistically significant, I used a parametric bootstrap to generate the null distribution of $\Delta$AIC values ($\Delta$AIC is the difference in AIC between competing models). Specifically, I sampled 1000 random datasets from the estimated model with main effects of Raunki\ae r life form and \el~but no interaction. I fit these simulated datasets to models with and without interactions and  calculated $\Delta$AIC. The statistical significance of the observed $\Delta$AIC is the proportion of simulated $\Delta$AIC greater than the observed.

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}). When classified based on plant habit, growth form interacted with \el less ($\Delta \textrm{AIC} = \Sexpr{round(AIC(fitSR_hf_g) - AIC(fitSR_hf_f), 1)}$; Fig.~\ref{fig:SR-hf}). Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore we focus hereafter on those analyses. Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SR-lf}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el.

Although there was significant phylogenetic signal in the residual variation of stomatal ratio (see R code), the total variation among these species was consistent with a trait at stationarity. Specifically, the simulated residual trait variation, after accounting for Raunki\ae r life form and \el, from the actual tree was not significantly greater than that simulated from a tree where traits had reached stationarity (paired $t$-test, $P = \Sexpr{pval2latex(ttest$p.value)}$). Hence, phylogenetic nonindependence is an important statistical consideration, but phylogeny does not explain stomatal trait variation among British angiosperms.

@@

Among British angiosperms, phylogenetic comparative analyses suggest that selection favors amphistomy in high light habitats among fast-growing plants, but not shrubs and trees. This is a significant advance over previous studies that considered each factor in isolation and/or did not use modern approaches to control for phylogenetic nonindenpendence. For example, pioneering studies by \cite{Salisbury_1927} first suggested that amphistomy is associated with herbs in open habitats, albeit without formal statistical tools to disentangle these effects. Later work by \cite{Peat_Fitter_1994b} demonstrated these trends again using family-level comparisons, a basic method to account for phylogenetic nonindependence \citep[see also][]{Mott_etal_1982, Beerling_Kelly_1996}. However, this approach is still problematic because traits like growth from can be highly phylogenetically conserved. For example, orders like Fagales contain multiple families dominated by hypostomatous trees, hence it is premature to conclude that this correlation is biologically meaningful without properly accounting for phylogenetic nonidendence. By combining trait, ecological, and phylogenetic datasets on British angiosperms, we now know that not only do both light and growth form influence stomatal ratio, but in fact their effects may reinforce one another. Based on information criteria, light may be a more important factor than growth form or their interaction (Table~\ref{table:srAIC}), consistent with previous studies indicating a dominant role o flight \citep{Mott_etal_1982, Jordan_etal_2014, Bucher_etal_2017}.

The interaction between light and growth form among British angiosperms suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesis (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs). This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}. Along these lines, Raunki\ae r lifeform may explain stomatal ratio better than plant habit (Table~\ref{table:srAIC}) because it is a better proxy for life history characteristics. For example, on an axis of `fast' to `slow' life history, geophytes more closely resemble phanerophytes than do chamaephytes or hemicryptophytes \citep{Salguero-Gomez_etal_2016}. Similarly, the relationship between light and stomatal ratio for geophytes was intermediate between that for phanerophytes and chamaephytes/hemicryptophytes (Fig.~\ref{fig:hf-light}). These comparisons indirectly suggest that both high light and fast life history are necessary to induce strong selection for amphistomy. The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g., containing both therophyte/annual and perennial forms). I predict that amphistomy will be favored more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light, and much more strongly than a perennial grown in low light. Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991, Fontana_etal_2017}.

@@

The weak or nonsignificant relationship between \el~and stomatal ratio in geophytes and phanerophytes suggests that in some cases amphistomy may not reliably indicate open habitat without further information. For example, perhaps amphistomatous geophytes from partially shaded habitats are spring ephemerals, so they  experience high light during their growth phase, but this has not been tested. Likewise, phanerophytes (mostly tall trees) are almost always hypostomatous \citep[see also][]{Muir_2015}. Most British phanerophytes are tall, hypostomatous trees, but the exceptions are telling. For example, the most amphistomatous phanerophyte in this dataset is \textit{Brassica oleracea}, a short-statured biennial that has more in common physiologically with hemicryptophytes than other phanerophytes. The other amphistomatous phanerophytes in this data set (\textit{Populus nigra} and \textit{Lavatera arborea}) are fast-growing pioneer species.

### 2017-10-14 ms/ms.tex
Added analysis of phylogenetic variance

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}). When classified based on plant habit, growth form interacted with \el less ($\Delta \textrm{AIC} = 2.4$; Fig.~\ref{fig:SR-hf}). Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore we focus hereafter on those analyses. Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SR-lf}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. % Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).

@@

Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio. The contributions of adaxial density, abaxial density, and their covariance are 1.14, 0.38, and -0.53, respectively. This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.

Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}). Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 6.1 $\times10^{-7}$ and 2.9 $\times10^{-5}$, respectively). However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2.1$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.21 versus 0.1). Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.23-fold change in adaxial stomatal density versus a 1.1-fold change in abaxial stomatal density. The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $1.7 $\times10^{-11}$). These results together imply that total stomatal density increases with \el, but the response is mediated mostly by increases in adaxial stomatal density.

@@

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform predict stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ Raunki\ae r lifeform
  & 0.46
  & 0.068
  & 0.34
  & 12
  & -33.3
  & 90.6
  & 0 \\
    \el~$\times$ growth form
  & 0.46
  & 0.07
  & 0.32
  & 12
  & -38.2
  & 100.3
  & 9.8 \\
    \el~+ Raunki\ae r lifeform
  & 0.47
  & 0.072
  & 0.32
  & 8
  & -40.3
  & 96.5
  & 6 \\
     \el~+ growth form
  & 0.51
  & 0.08
  & 0.31
  & 8
  & -43.3
  & 102.7
  & 12.1 \\
    Raunki\ae r lifeform
  & 0.34
  & 0.067
  & 0.15
  & 7
  & -79.2
  & 172.4
  & 81.8 \\
    growth form
  & 0.35
  & 0.069
  & 0.13
  & 7
  & -82.5
  & 178.9
  & 88.4 \\
    \el
  & 0.64
  & 0.108
  & 0.26
  & 4
  & -59.3
  & 126.6
  & 36.1 \\
    null
  & 0.29
  & 0.067
  & 0
  & 3
  & -107.5
  & 221
  & 130.5 \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}

>>>>

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}). When classified based on plant habit, growth form interacted with \el less ($\Delta \textrm{AIC} = 2.4$; Fig.~\ref{fig:SR-hf}). Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore we focus hereafter on those analyses. Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SR-lf}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el.

Although there was significant phylogenetic signal in the residual variation of stomatal ratio (see R code), the total variation among these species was consistent with a trait at stationarity. Specifically, the simulated residual trait variation, after accounting for Raunki\ae r life form and \el, from the actual tree was not significantly greater than that simulated from a tree where traits had reached stationarity (paired $t$-test, $P = 0.331$). Hence, phylogenetic nonindependence is an important statistical consideration, but phylogeny does not explain stomatal trait variation among British angiosperms.

@@

Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio. The contributions of adaxial density, abaxial density, and their covariance are 1.12, 0.38, and -0.5, respectively. This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.

Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}). Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 1.2 $\times10^{-8}$ and 8.9 $\times10^{-7}$, respectively). However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.24 versus 0.12). Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.27-fold change in adaxial stomatal density versus a 1.13-fold change in abaxial stomatal density. The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $2.5 $\times10^{-10}$). These results together imply that total stomatal density increases with \el, but the response is mediated mostly by increases in adaxial stomatal density.

@@

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform predict stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ Raunki\ae r lifeform
  & 0.46
  & 0.068
  & 0.34
  & 12
  & -33.2
  & 90.4
  & 0 \\
    \el~$\times$ growth form
  & 0.46
  & 0.07
  & 0.32
  & 12
  & -38.2
  & 100.4
  & 9.9 \\
    \el~+ Raunki\ae r lifeform
  & 0.46
  & 0.071
  & 0.32
  & 8
  & -40.2
  & 96.4
  & 6 \\
     \el~+ growth form
  & 0.51
  & 0.08
  & 0.31
  & 8
  & -43.4
  & 102.7
  & 12.3 \\
    Raunki\ae r lifeform
  & 0.34
  & 0.067
  & 0.15
  & 7
  & -79.2
  & 172.4
  & 82 \\
    growth form
  & 0.35
  & 0.069
  & 0.13
  & 7
  & -82.5
  & 179.1
  & 88.6 \\
    \el
  & 0.64
  & 0.107
  & 0.26
  & 4
  & -59.3
  & 126.6
  & 36.2 \\
    null
  & 0.29
  & 0.067
  & 0
  & 3
  & -107.6
  & 221.1
  & 130.7 \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}
